Abstractions for Multi-Sorted Substitutions

### 2024-03-13 paper/paper.tex
Update.

  As we defined the \ACode{\ACon{Kit}} structure before the
  \ACode{\ACon{Traversal}} structure, the fields of
  \ACode{\ACon{Kit}} could not use map application \ACode{\AField{\_⋯\_}} in
  their types.
  This prevented us to include another useful axiom into the
  \ACode{\ACon{Kit}} structure. As this axiom also needs to be proved
  separately for variables and terms, we define a new structure for
  it which extends a \ACode{\ACon{Kit}}:
  \LibCode\KWkKit

  The \ACode{\AField{wk-`/id}} axiom
  explains the \ACode{\AField{wk}} function by how it acts on \emph{terms}.
  It is the counterpiece to the \ACode{\ACon{Kit}} axiom
  \ACode{\AField{wk-id/`}}, which explains the \ACode{\AField{wk}}
  function by how it acts on \emph{variables}.

  This lemma is useful for proving extensional equalities between maps
  involving weakening, where \ACode{\AField{`/id-injective}} allows to
  add \ACode{\AField{`/id}} on both sides of the equation, such that
  \AField{wk-`/id} can be used to make further progress.

  The instantiations of the \ACode{\ACon{WkKit}} are straightforward:
  \LibCode\KWkKitInstances

@@

  Previous work on kits\cite{DBLP:journals/jar/BentonHKM12} addresses this
  issue by duplicating the definitions and using tactics to reduce
  boilerplate in proofs.
  In contrast, we define structures similar to \ACode{\ACon{Kit}} and
  \ACode{\ACon{Traversal}}, which allow us to abstract over all four
  compositions and use the same trick as before to eliminate the
  dependencies.
  % TODO
  While composition can be derived, our approach has the benefit, that
  we have gained a new level of generality, which enables us in the next
  sections to unify type preservation of all substitution and renaming
  operations into one lemma.

@@

  The definitions reveal two interesting properties:
  \begin{enumerate}
  \item
    If we compose two maps \ACode{ϕ₁} and \ACode{ϕ₂}, then the
    resulting map is a renaming, iff both \ACode{ϕ₁} and \ACode{ϕ₂} are
    renamings.
    In other words:
    if \ACode{ϕ₁} is a \ACode{K₁}-map and \ACode{ϕ₂} is a \ACode{K₂}-map,
    then the result is a \ACode{(K₁ ⊔ K₂)}-map, where \ACode{⊔} refers to the
    lattice for \ACode{\{ \ADef{Kᵣ} , \ADef{Kₛ} \}}
    generated by \ACode{\ADef{Kᵣ} < \ADef{Kₛ}}.
  \item
    All four compositions first apply \ACode{ϕ₁} to \ACode{x},
    and then apply \ACode{ϕ₂} to the result.
    If \ACode{ϕ₁} is a renaming, then this result is another variable, but
    if \ACode{ϕ₁} is a substitution, then this result is a term.
  \end{enumerate}
  With the \ACode{\ACon{Kit}} abstraction, we can easily abstract over
  \ACode{ϕ₂} being a renaming or a substitution:

@@

  But to abstract over \ACode{ϕ₁}, the \ACode{\ACon{Kit}} abstraction
  is not sufficient: while it allows us to abstract over what we are applying, i.e.
  a renaming or a substitution, it does not allow us to
  abstract over what we are applying it to, i.e. a variable or a term.
  For the latter we have two distinct operations \ACode{\ADef{\_\&\_}}
  and \ACode{\ADef{\_⋯\_}}.

  To fill this gap, we introduce a new abstraction that we call a
  \ACode{\ACon{CKit}},
  which provides an operation \ACode{\AField{\_\&/⋯\_}} that unifies
  \ACode{\ADef{\_\&\_}} and \ACode{\ADef{\_⋯\_}}. This allows us to
  define a general composition, which informally can be described as:
  \ExampleCode\KComposition

@@

  The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
  \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
  subsequent lemmas to make use of the lemmas we previously proved
  about \ACode{\AField{\_⋯\_}}.

>>>>

  As we defined the \ACode{\ACon{Kit}} structure before the
  \ACode{\ACon{Traversal}} structure, the fields of
  \ACode{\ACon{Kit}} could not use map application \ACode{\AField{\_⋯\_}} in
  their types.
  This prevented us to include another useful axiom into the
  \ACode{\ACon{Kit}} structure. As this axiom also needs to be proved
  separately for variables and terms, we define a new structure for
  it which extends a \ACode{\ACon{Kit}}:
  \LibCode\KWkKit
  \noindent The \ACode{\AField{wk-`/id}} axiom
  explains the \ACode{\AField{wk}} function by how it acts on \emph{terms}.
  It is the counterpiece to the \ACode{\ACon{Kit}} axiom
  \ACode{\AField{wk-id/`}}, which explains the \ACode{\AField{wk}}
  function by how it acts on \emph{variables}.
  %
  This lemma is useful for proving extensional equalities between maps
  involving weakening, where \ACode{\AField{`/id-injective}} allows to
  add \ACode{\AField{`/id}} on both sides of the equation, such that
  \AField{wk-`/id} can be used to make further progress.
  %
  The instantiations of the \ACode{\ACon{WkKit}} are straightforward:
  \LibCode\KWkKitInstances

@@

  Previous work on kits\cite{DBLP:journals/jar/BentonHKM12} addresses this
  issue by duplicating the definitions and using tactics to reduce
  boilerplate in proofs.
  In contrast, we define structures similar to \ACode{\ACon{Kit}} and
  \ACode{\ACon{Traversal}}, which allow us to abstract over all four
  compositions and use the same trick as before to eliminate the
  dependencies.
  With the help of a general map composition, we can prove lemmas
  about the interactions of general maps, which is crucial for proving a
  type preservation lemma for general map application instead of individual
  lemmas for renamings and substitutions.

@@

  The definitions reveal two interesting properties:
  \begin{enumerate}
  \item
    If we compose two maps \ACode{ϕ₁} and \ACode{ϕ₂}, then the
    resulting map is a renaming, iff both \ACode{ϕ₁} and \ACode{ϕ₂} are
    renamings.
    In other words:
    if \ACode{ϕ₁} is a \ACode{K₁}-map and \ACode{ϕ₂} is a \ACode{K₂}-map,
    then the result is a \ACode{(K₁ ⊔ K₂)}-map, where \ACode{⊔} refers to the
    lattice for \ACode{\{ \ADef{Kᵣ} , \ADef{Kₛ} \}}
    generated by \ACode{\ADef{Kᵣ} < \ADef{Kₛ}}.
  \item
    All four compositions first apply \ACode{ϕ₁} to \ACode{x},
    and then apply \ACode{ϕ₂} to the result.
    If \ACode{ϕ₁} is a renaming, this result is another variable, but
    if \ACode{ϕ₁} is a substitution, this result is a term.
  \end{enumerate}
  With the \ACode{\ACon{Kit}} abstraction, we can easily abstract over
  \ACode{ϕ₂} being a renaming or a substitution:

@@

  \noindent But to abstract over \ACode{ϕ₁}, the \ACode{\ACon{Kit}} abstraction
  is not sufficient: while it allows us to abstract over what we are applying, i.e.
  a renaming or a substitution, it does not allow us to
  abstract over what we are applying it to, i.e. a variable or a term.
  For the latter we have two distinct operations \ACode{\ADef{\_\&\_}}
  and \ACode{\ADef{\_⋯\_}}.

  To fill this gap, we introduce a new abstraction that we call a
  \emph{compose kit} (\ACode{\ACon{CKit}}),
  which provides an operation \ACode{\AField{\_\&/⋯\_}} that unifies
  \ACode{\ADef{\_\&\_}} and \ACode{\ADef{\_⋯\_}}. This allows us to
  define a general composition as follows:
  \ExampleCode\KComposition

@@

  The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
  \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
  subsequent lemmas to make use of the lemmas that we have already proved
  for \ACode{\AField{\_⋯\_}}.
